---
title: "US Youth Smoking Analysis"
subtitle: "Which group of youth should be the target of smoking prevention?"
author: 
  - Huakun Shen
thanks: "Code and data are available at: https://github.com/HuakunShen/NYTS-Analysis"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "An statistical analysis has been conducted to analyze national youth tobacco usage using NYTS survey prodiced by CDC. The use of tobacco is known to be addictive and harmful to human health, thus preventing tobacco consumption during the youth is crucial, as youth is likely the time people first start to consume tobacco. It's important to know who are more likely to smoke in order to target this group of people and apply prevention means. It is found that Males high school students whose race is Hispanic, White and Black (in decreasing order) are more likely to smoke. Curiosity is also an important factor. Keeping students in a healthy environment is important."
output:
  bookdown::pdf_document2
toc: FALSE
keywords: 
  - smoking
  - youth
  - US
bibliography: references.bib
header-includes:
  - \usepackage{subfig}
  - \usepackage{float}
  - \floatplacement{figure}{H}
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(kableExtra)
library(here)
library(gridExtra)
# source(here::here("scripts/preprocess.R"))
df2015 <- read.csv(here::here("inputs/data/df2015.csv"))
df2016 <- read.csv(here::here("inputs/data/df2016.csv"))
df2017 <- read.csv(here::here("inputs/data/df2017.csv"))
df2018 <- read.csv(here::here("inputs/data/df2018.csv"))
df2019 <- read.csv(here::here("inputs/data/df2019.csv"))
df2020 <- read.csv(here::here("inputs/data/df2020.csv"))
```

# Introduction

The use of tobacco, cigarette and other products related to smoking is known to be addictive 
and have a negative effect on humans' health. As many habits are formed during the youth, it may be more effective to intervene during this period.
Many factors can affect humans' habits. To effectively influence the youth, we should know which group of people we are targeting (are more likely to smoke). For example, what's the relationship between race, age, sex and smoking.
I am also interested in the trend of the percentage of smoking population, so that we can have an idea of whether the smoking prevention means are effective.

After doing the analysis I found that high school males students whose race is Hispanic, white or black are more likely to smoke. 
Curiosity is also a big factor, lowing students' curiosity could also potentially prevent smoking. 
However, there is a paradox about curiosity. 
Teaching students about the risks of smoking could raise their curiosity, not teaching the risk is obviously not improper.
So, providing a healthy environment for students and teaching them about the risk of tobacco consumption are important.

# Data

This analysis is done using R [@citeR], dplyr[@dplyr], tidyverse[@tidyverse], here[@here]. Graphs and tables are generated using and ggplot2[@ggplot2], kableExtra[@kableExtra], and gridExtra[@gridExtra].

## Data Source

The data used in this report was retrieved from NYTS[@NYTS] from CDC. 
The dataset was designed to provide national data on indicators key to the design, implementation and evaluation of comprehensive tobacco prevention and control programs. The goal of the dataset is to reduce tobacco use among youth. Historical data is available from 1999 to 2021 (some years are skipped).
In this report, I used the data of recent years (2015-2020) to see the trend of tobacco use in the youth in the US.

The data collected in different years are different. There are hundreds of columns for each year. 
So I pick some of the common columns that are representative, they are

- race: Asian, Black, White, Hispanic, AI/AN or NHOPI
- age: age with a offset of -9, i.e. 1 if a person is 9 years old, 2 if 10 years old and so on
- schooltype: middle school or high school
- smoked: whether have been smoked in the past 30 days
- curious: have been curious about smoking a cigarette
  - Definitely yes and probably yes are both considered **TRUE**
  - Definitely no and probably no are both considered **FALSE**

Since the data raw dataset's columns are not labelled in English, I pre-processed the data to make it easier to understand.

The **smoked** column is a combination of multiple survey questions (using **or** operator), the description is as follows.
Which means, if either of the following columns are **TRUE**, **smoked** will be TRUE.

| Col | Description |
| --- | --- |
| ccigt_r | Smoked cigarettes on 1 or more days in the past 30 days |
| ccigar_r | Smoked cigars, cigarillos, or little cigars on 1 or more days in the past 30 days |
| cslt_r | Used chewing tobacco, snuff, or dip on 1 or more days in the past 30 days |
| cpipe_r | Smoked tobacco out of a pipe (not a hookah or waterpipe) on 1 or more days in the past 30 days |
| cbidis_r | Smoked bidis on 1 or more days in the past 30 days |
| chookah_r | Smoked tobacco out of a hookah or waterpipe on 1 or more days in the past 30 days |
| csnus_r | Used snus, such as Camel or Marlboro Snus, on 1 or more days in the past 30 days |
| cdissolv_r | Used dissolvable tobacco, such as Ariva, Camel orbs, Camel sticks, Camel strips, or Stonewall, on 1 or more |
| celcigt_r | Used electronic cigarettes, such as Ruyan or NJOY, on 1 or more days in the past 30 days |

Basically, all kinds of tobacco use in the past 30 days are considered as **smoked**.

This dataset/survey is chosen because it's published by CDC (Centers for Disease Control and Prevention), which should be the most official and authoritative institution for this type of data in the United States. CDC has two other surveys, they are "Alaska Native Adult Tobacco Survey Guidance Manual" and "American Indian Adult Tobacco Survey" which are for 2 minority groups that can not represent the US. A report from WHO called "WHO Global Tobacco Control Report 2011 - Data" has a 404 not found page. NYTS is the only dataset I found that contains data I need and also historical data for many years.

Figure \@ref(fig:dataset-preview) shows a preview of the dataset (selected properties, 2015) by displaying the distribution of smoking population (not percentage) by various properties. The datasets for other years have exactly the same structure.

```{r, dataset-preview, fig.cap="Smoking Population Distribution", echo=FALSE}
# Plot some frequency plots for different attributes
smoked2015 <- df2015 %>% filter(smoked==1) %>% filter(!is.na(Sex))
sex_hist <- ggplot(smoked2015, aes(Sex)) + geom_bar(fill = "#0073C2FF") + ggtitle("Population Distribution By Sex (2015)") + theme(plot.title = element_text(size=10))
smoked2015 <- df2015 %>% filter(smoked==1) %>% filter(!is.na(Age))
smoked2015$RealAge = smoked2015$Age + 9
age_hist <- ggplot(smoked2015, aes(RealAge)) + geom_bar(fill = "#0073C2FF") + ggtitle("Population Distribution By Age (2015)") + xlab("Age") + theme(plot.title = element_text(size=10))
smoked2015 <- df2015 %>% filter(smoked==1) %>% filter(!is.na(race))
race_hist <- ggplot(smoked2015, aes(race)) + geom_bar(fill = "#0073C2FF") + ggtitle("Population Distribution By Race (2015)") + xlab("Race") + theme(plot.title = element_text(size=10))
smoked2015 <- df2015 %>% filter(smoked==1) %>% filter(!is.na(schooltype))
schooltype_hist <- ggplot(smoked2015, aes(schooltype)) + geom_bar(fill = "#0073C2FF") + ggtitle("Population Distribution By School Type (2015)") + xlab("School Type") + theme(plot.title = element_text(size=8))
smoked2015 <- df2015 %>% filter(!is.na(smoked))
smoked_hist <- ggplot(smoked2015, aes(smoked)) + geom_bar(fill = "#0073C2FF") + ggtitle("Population Distribution By Smoked (2015)") + xlab("Smoked") + theme(plot.title = element_text(size=10))
grid.arrange(sex_hist, age_hist, race_hist, schooltype_hist, smoked_hist, nrow = 3)
```


# Model

```{r model-definition, echo=FALSE}
# define the model
model2015 <- glm(smoked ~ Age + Sex + race + schooltype + curious, data=df2015 %>% filter(race!="AI/AN") %>% filter(race!="NHOPI"), family=binomial(link = "logit"))
model2016 <- glm(smoked ~ Age + Sex + race + schooltype + curious, data=df2016 %>% filter(race!="AI/AN") %>% filter(race!="NHOPI"), family=binomial(link = "logit"))
model2017 <- glm(smoked ~ Age + Sex + race + schooltype + curious, data=df2017 %>% filter(race!="AI/AN") %>% filter(race!="NHOPI"), family=binomial(link = "logit"))
model2018 <- glm(smoked ~ Age + Sex + race + schooltype + curious, data=df2018 %>% filter(race!="AI/AN") %>% filter(race!="NHOPI"), family=binomial(link = "logit"))
model2019 <- glm(smoked ~ Age + Sex + race + schooltype + curious, data=df2019 %>% filter(Sex != "N") %>% filter(race!="AI/AN") %>% filter(race!="NHOPI"), family=binomial(link = "logit"))
model2020 <- glm(smoked ~ Age + Sex + race + schooltype, data=df2020 %>% filter(race!="AI/AN") %>% filter(race!="NHOPI"), family=binomial(link = "logit"))
```

$$Y\sim Bernoulli(\pi)$$
$$\log \frac{\pi}{1-\pi}=X\beta$$
The response variable $Y$ is **smoked** boolean variable.

The predictor variables include

- Age
- Sex
- Race
- School Type
- curious

The type of model used is GLM (Generalized Linear Model), logistic regression.

Logistic regression is used because the response variable **smoked** has a binary value.

See [Model Summary] section for description about the intercept/baseline group because this is determined by R [@citeR].

A model card is included in the appendix, information are obtained from [@penn-GLM].

# Results

## Model Summary

In the following model summaries, the intercept/baseline is the same.
The baseline group is Female, Age=9, Asian, high school students who have never been curious about smoking a cigarette. 

Since the sample size of AI/AN and NHOPI are too low, these 2 races are removed from the models (so that they don't serve as the baseline group).

Most selected variables are significant except for **schooltype**, which has been significant in 2 of the 5 years.

Overall, Asians, Females, middle school (low-age) students who had less curiosity in smoking are less likely to smoke.

### Model 2015

```{r, summary-2015, echo=FALSE}
# display model summary coefficients for 2015
knitr::kable(summary(model2015)$coef, digits = 4)
```

All variables are significant except for **schooltype**.

### Model 2016

```{r, summary-2016, echo=FALSE}
# display model summary coefficients for 2016
knitr::kable(summary(model2016)$coef, digits = 4)
```

All variables are significant except for **schooltype**.

### Model 2017

```{r, summary-2017, echo=FALSE}
# display model summary coefficients for 2017
knitr::kable(summary(model2017)$coef, digits = 4)
```

All variables are significant.

### Model 2018

```{r, summary-2018, echo=FALSE}
# display model summary coefficients for 2018
knitr::kable(summary(model2018)$coef, digits = 4)
```

All variables are significant.

### Model 2019

```{r, summary-2019, echo=FALSE}
# display model summary coefficients for 2019
knitr::kable(summary(model2019)$coef, digits = 4)
```

All variables are significant except for **schooltype**.

### Model 2020

```{r, summary-2020, echo=FALSE}
# display model summary coefficients for 2020
knitr::kable(summary(model2015)$coef, digits = 4)
```

All variables are significant.

## Smoking Population Trend

```{r trend-plot, fig.cap="Trend of Sample Size and Smoker Population", echo=FALSE, fig.height=2}
# plot a trend line plot for smoking population and sample size from 2015 to 2020
years <- c(2015, 2016, 2017, 2018, 2019, 2020)

# calculate smoking percentage
percentages <- c(
  (df2015 %>% filter(smoked == 1) %>% count / df2015 %>% filter(!is.na(smoked)) %>% count)$n,
  (df2016 %>% filter(smoked == 1) %>% count / df2016 %>% filter(!is.na(smoked)) %>% count)$n,
  (df2017 %>% filter(smoked == 1) %>% count / df2017 %>% filter(!is.na(smoked)) %>% count)$n,
  (df2018 %>% filter(smoked == 1) %>% count / df2018 %>% filter(!is.na(smoked)) %>% count)$n,
  (df2019 %>% filter(smoked == 1) %>% count / df2019 %>% filter(!is.na(smoked)) %>% count)$n,
  (df2020 %>% filter(smoked == 1) %>% count / df2020 %>% filter(!is.na(smoked)) %>% count)$n
)
trend_df <- data.frame(year=years, percentage=percentages)


smoker_percentage_trend <- trend_df %>% ggplot(aes(x=year, y=percentages)) +
  geom_line(color="red")+
  geom_point() + ggtitle("Percentage Of Smoker Trend")

# calculate sample size
counts <- c(
  (df2015 %>% count)$n,
  (df2016 %>% count)$n,
  (df2017 %>% count)$n,
  (df2018 %>% count)$n,
  (df2019 %>% count)$n,
  (df2020 %>% count)$n
)
trend_df <- data.frame(year=years, count=counts)
sample_size_trend <- trend_df %>% ggplot(aes(x=year, y=count)) +
  geom_line(color="red")+
  geom_point() + ggtitle("Sample Size Trend")
grid.arrange(sample_size_trend, smoker_percentage_trend, nrow = 1)
```

From figure \ref(fig:trend-plot), we can see a large decrease in smoker percentage from 2015 to 2016, then a large increase in 2018 and 2019.
There is a huge decrease in smoking percentage in 2020 with a big decrease in sample size in 2020 as well.

## Density

```{r density-plot, fig.cap="Smoker Density By Group", fig.height=8, echo=FALSE}
# plot density and frequency plots grouping by different properties and "smoked"
# sex density plots
sex_density <- df2015 %>% filter(!is.na(smoked)) %>% filter(!is.na(Sex)) %>% ggplot(aes(Sex)) + geom_bar(aes(fill=smoked)) + ggtitle("Sex Frequency")

# race plots
race_df <- df2015 %>% filter(!is.na(smoked)) %>% filter(!is.na(race))
race_count_df <- merge(race_df %>% group_by(race, smoked) %>% count %>% rename(count = n), race_df %>% group_by(race) %>% count %>% rename(total = n), by="race")
race_count_df$percentage <- race_count_df$count / race_count_df$total
race_density <- ggplot(race_count_df, aes(fill=smoked, y=percentage, x=race)) + 
    geom_bar(position="fill", stat="identity") + theme(axis.text.x = element_text(angle = 30, vjust = 0.5, hjust=1)) + xlab("Race") + ggtitle("Race Density")

race_freq <- df2015 %>% filter(!is.na(smoked)) %>% filter(!is.na(race)) %>% ggplot(aes(race)) + geom_bar(aes(fill=smoked)) + theme(axis.text.x = element_text(angle = 30, vjust = 0.5, hjust=1)) + xlab("Race") + ggtitle("Race Frequency")


# remove AI/AN and NHOPI from race because their sample size is too small and could influence the plot scale
race2_df <- df2015 %>% filter(!is.na(smoked)) %>% filter(!is.na(race)) %>% filter(race!="AI/AN") %>% filter(race!="NHOPI")
race_count_df <- merge(race2_df %>% group_by(race, smoked) %>% count %>% rename(count = n), race2_df %>% group_by(race) %>% count %>% rename(total = n), by="race")
race_count_df$percentage <- race_count_df$count / race_count_df$total

race_density_2 <- ggplot(race_count_df, aes(fill=smoked, y=percentage, x=race)) + 
    geom_bar(position="fill", stat="identity") + theme(axis.text.x = element_text(angle = 30, vjust = 0.5, hjust=1)) + xlab("Race") + ggtitle("Race Density")

race_freq_2 <- df2015 %>% filter(!is.na(smoked)) %>% filter(!is.na(race)) %>% ggplot(aes(race)) + geom_bar(aes(fill=smoked)) + theme(axis.text.x = element_text(angle = 30, vjust = 0.5, hjust=1)) + xlab("Race") + ggtitle("Race Frequency")

# Age plots
df2015$RealAge <- df2015$Age + 9
RealAge_df <- df2015 %>% filter(!is.na(smoked)) %>% filter(!is.na(RealAge))
RealAge_count_df <- merge(RealAge_df %>% group_by(RealAge, smoked) %>% count %>% rename(count = n), RealAge_df %>% group_by(RealAge) %>% count %>% rename(total = n), by="RealAge")
RealAge_count_df$percentage <- RealAge_count_df$count / RealAge_count_df$total
age_density <- ggplot(RealAge_count_df, aes(fill=smoked, y=percentage, x=RealAge)) + 
    geom_bar(position="fill", stat="identity") + theme(axis.text.x = element_text(angle = 30, vjust = 0.5, hjust=1)) + xlab("Age") + ggtitle("Age Density")

age_freq <- df2015 %>% filter(!is.na(smoked)) %>% filter(!is.na(RealAge)) %>% ggplot(aes(RealAge)) + geom_bar(aes(fill=smoked)) + xlab("Age") + ggtitle("Age Frequency") + ggtitle("Age Frequency")

# school type plots
school_type_density <- df2015 %>% filter(!is.na(smoked)) %>% filter(!is.na(schooltype)) %>% ggplot(aes(schooltype)) + geom_bar(aes(fill=smoked)) + xlab("School Type") + ggtitle("School Type Frequency")
grid.arrange(race_density, race_freq, race_density_2, race_freq_2, age_density, age_freq, school_type_density, sex_density, nrow = 4)
```

```{r smoker-percentage-by-race, echo=FALSE}
# make a table showing the percentage of smokers for each race
race_count_df$Percentage = round(race_count_df$percentage * 100, 2)
df <- race_count_df %>% rename(Race=race) %>% filter(smoked==TRUE)
knitr::kable(df[c("Race", "Percentage")], caption="Smoker Percentage By Race")
```


Figure \@ref(fig:density-plot) plots the density of smoked population by different groups.
The general trend is
1. Higher percentage of males smokes than females
2. As age increases, more adolescents tend to smoke (higher percentage of high school students smoke than middle school)
  1. When age is 10-11, the smoking ratio is super high, but this is probably because the sample size is too small. See Age Frequency plot.
3. Higher percentage of AI/AN and NHOPI smoke than other races (but their sample size is very small)
  1. See table \@ref(tab:smoker-percentage-by-race). Hispanic has the highest percentage of smokers, then White, Black and Asian
  


# Discussion

## Trend

From figure \@ref(fig:trend-plot), we can see an overall increasing trend of smoking youth population.
The drop of sample size in 2020 could be due to Covid 19, and this could have an impact on the accuracy of collected results.
However, even with the large drop in sample size in 2020, the sample size is still pretty large.
So, another hypothesis is that smokers has been reduced due to Covid 19 because more people tend to stay home.
I believe the overall trend is still an increasing trend with the data of recent years. It may be more accurate if we can look at the historical data of the years before 2015.

## Race

Figure \@ref(fig:density-plot) shows that, among the 4 races with the largest populations (Asian, Black, Hispanic and White), Asian has the lowest percentage of smokers.
The other 3 groups have relative high smoking percentage.

This hypothesis is also supported by GLM. From [Model Summary] results, we can see that all variables are significant based on the p-values. 
Asian is the baseline group, raceBlack, raceWhite and raceHispanic all have positive estimate value.

## Other Variables

In [Model Summary] section we can see that **SexM** variable is significant and positive for all years.
Meaning that more male tends to smoke than female does.

Age is positive, meaning that as age increases, youths are more likely start to smoke.

**schooltypeMiddle** significant for 2 of the 5 years. It is negative meaning that high school students are more likely to smoke than middle school students are,
which is expected as **schooltype** is directly related to age.

**curious** is also always significant and positive, meaning that students who have been curious in smoking are more likely to smoke later.

## Conclusion

With the evidence from statistical models and plots, we can conclude the target for smoking prevention.
The main target are high school male students whose race is hispanic, white or black.

This doesn't mean middle school students should not be focused on. If middle school students start to smoke, 
it's more likely for them to smoke later or even influence their classmates. This is evidenced by the **curious** variable.
As more classmates/friends are smoking, a student should be curious about smoking and are more likely to smoke eventually.

## Limitation

The dataset has a serious limitation when we need to perform an analysis across multiple years.
The number of survey questions is increasing every year, more columns are added to the dataset, which is fine.
What's troublesome is that many column names are not consistent across years, some questions are even missing in later years.
Which could cause serious mistake if not realized, and takes lots of time to find the column names and match them. 


\appendix

# Appendix {-}

```{r, model-card, out.width="\\linewidth", include=TRUE, fig.align="center", fig.cap=c("GLM Model Card"), echo=FALSE}
knitr::include_graphics(here::here("outputs/GLM-model-card.pdf"))
```



\newpage


# References


